Computing likelihoods under the SMC for a general class of ARGs.

### 2024-07-11 paper.tex

Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of ARGs, 
the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, that is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
We require a single half-open interval to describe
the ancestral
material associated with each common ancestor node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is $L(t)$,
the set of labeled lineages extant at time $t$,
where each lineage $X_j(t)$ is labeled by an integer $j$
and represented by a half-open interval delimiting the
ancestral material. Any lineage can thus be summarized as $X_j(t) = [x_{j}, y_{j})$,
for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards in time, $t=0$ is today, and 
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit of genomic length and time:
so, if $T(t) = \sum_i |X_j(t)|$ is the sum of the total sizes of all lineages,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
given new, unique labels and added to $L(t)$
(and lineage $j$ is removed).
This operation keeps the total amount of ancestral material unchanged.

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.

Although coalescence is reciprocal, here we'll define a strict total order
on all lineages in $L(t)$ based on their leftmost starting points (and index to break ties).
For any such strict total order,
the instantaneous rate of coalescence then equals $\lambda$ multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$. For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e., 
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify the likelihood computations. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

[Do we need to get into this above-the-root subtelty here?]
To ensure that the ancestral material carried by a single lineage can at
all time be represented by a single half-open interval, we only update the
endpoints of the involved lineages following common ancestry or recombination events
such that only the minimal half-open interval whose endpoints have not fully coalesced
is used to update $L(t)$. Note that empty intervals are not added to the state.
%By reasoning about the process in this way we can keep representing 
% all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% and the process terminates.

%After any coalescence or recombination event,
%the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
%are replaced by a single segment $[x, y)$ whose endpoints have not fully
%coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
%and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
%are not added back to $L(t)$.

Finally, by describing the SMC
backwards in time, some intervals contained within the half-open
interval representing a lineage might have fully coalesced before all the genetic
material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
intervals however do not affect the marginal genealogies.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    \caption{
        Graphical representation of the events associated with the edge $(p,c,x,y)$.
        The lineage carried by $c$ spans $[x_c,y)$ until a recombination event (blue dotted line) splits it
        into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
        bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
        Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
        green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
        \label{fig:likelihood}
    }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
Under the SMC,
we therefore know there was no recombination between $x$ and $y$
on any of the lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occur, then the number of links not
affected by recombination is 1 less (see eq.~\eqref{eq:depth}).

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(c, p) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
One significant advantage of this formulation is that the computation time
is linear in the number of edges, and an algorithm can be devised to 
to efficiently keep track of $I_e(t)$ (see Appendix).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
We can also accommodate stepwise changes in the 
coalescence rate (or recombination rate)
through time by intersecting the internode intervals with the rate change
intervals; rate changes along the genome can be incorporated similarly.
Finally, we have only considered the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is given by
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(c, p) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.

>>>>

Linking the resulting tree-valued process along the genome back to
the relationship between the ``graph'' and
``local trees'' representation of ARGs, 
the Markovian property can be formulated as follows: 
the distribution of the next tree within the series
of local trees can be generated only knowing the current tree,
provided that we remove recombinant nodes entirely,
and erase the unary portions of coalescent nodes.
To see that the locally unary nodes make the process non-Markov,
consider that a node in a tree might be unary either because
it's about to be (in the left-to-right sense) a coalescent node 
or it already has been:
looking at the second tree in Fig.\ref{fig:smc-unary},
we know node $f$ will be coalescent further along the sequence
because it was not present in the first tree.
So, the first tree gives us information about subsequent trees
that the second tree alone could not.
Further note that by limiting common ancestry events to overlapping 
pairs of lineages, we are guaranteed that all nodes in the ARG 
only carry ancestral material, that is,
genetic material that is inherited by either one of a set of samples
\citep{wiuf_ancestry_1999}.
We therefore require a single half-open interval to describe
the ancestral
material associated with each node in the ARG.

We now define the process more formally, by reintroducing
the notation of \citet{mcvean_approximating_2005} required to describe
the likelihood computation. This section
does not deviate from the original description of the SMC apart from
how coalescable pairs are counted. This subtle nuance
greatly simplifies the ability to keep track of coalescable pairs going
backwards in time.
At any point in time, the state of the process is 
the set of labeled lineages extant at time $t$,
$L(t) = \{X_j(t)\}$.
A lineage $X_j(t) = [x_{j}, y_{j})$ is labeled by 
an integer $j$ and defines the 
half-open genomic interval of its ancestral material.
% Ea
% where each lineage $X_j(t)$ is labeled by an integer $j$
% and represented by a 
% ancestral material. 
% Any lineage can thus be summarized as 
% for some interval endpoints $x_{j} < y_{j}$.
Because we look backwards in time, $t=0$ is today, and 
$L(0)$ consists of $n$ sampled lineages, labelled $0$ to $n-1$,
each represented by a single interval spanning the entire genome.
Then, the process evolves by a succession of common ancestor and recombination
events until each segment of ancestral material is only present in one lineage.
The waiting time to next event is determined by these two competing processes
with exponentially distributed waiting times as outlined below.

Recombination is described by a Poisson process of rate $r$ per unit 
of genomic length and time:
so, if $T(t) = \sum_j |X_j(t)|$ is total length of ancestral material
carried by lineages at time $t$,
then recombination occurs at instantaneous rate $rT(t)$.
A recombination to the left of $x$
(i.e., between base pairs $x$ and $x-1$)
with $x_{j} < x <y_{j}$ splits lineage $j$ into two new lineages,
given new, unique labels. We then remove lineage $j$
and add the two new lineages to the state $L(t)$.
This operation keeps the total amount of ancestral material unchanged.

Assuming a randomly mating, diploid population of constant size $N_e$,
common ancestry occurs between
two \emph{overlapping} lineages at rate $\lambda = 1/(2N_e)$.
Here we deviate from the McVean and Cardin formulation,
and define a strict total order on all lineages in $L(t)$:
for each lineage $X_j(t) = [x_{j}, y_{j})$ we define its 
order by the left coordinate $x_j$ and label $j$ (to break ties).
The instantaneous rate of coalescence then equals $\lambda$ 
multiplied by the number of overlapping pairs,
i.e., $\lambda \sum_{i \neq j} I_{jk}$,
where
\begin{equation} \label{def:coal}
I_{jk} = \begin{cases}
    1 & X_j > X_k \text{ and } y_k > x_j \\
    0 & \text{otherwise.}
\end{cases}
\end{equation}
We say that lineage $k$ can coalesce \emph{into} lineage $j$ if $I_{jk} = 1$.
From the definition of the strict total order it follows that if $I_{jk} = 1$
then $I_{kj} = 0$. 
The newly formed lineage acquires the union of both intervals,
the original lineages are removed, and $L(t)$ is updated accordingly.
[JK notes: need to explain here that this deliberately introduced 
asymmetry doesn't affect the actual process. Also need to comment on 
how the process terminates. Basically has the same subtleties as McVean and 
Cardin, and not important for our purposes.]

For later use, also define $C_j(t)$ to be the set of
lineages that can coalesce into lineage $j$, i.e., 
$C_j(t) = \{X_k \in L(t) | I_{jk}(t) = 1\}$; 
thus, $|C_j(t)| = I_{j}(t) = \sum_{k} I_{jk}(t)$,
and the total rate of coalescence at time $t$ is $\sum_{j} |C_j(t)|$. 
Because $I_{jk}$ is defined in terms of a total order, the sets $C_j(t)$ are disjoint,
which will simplify likelihood computations later. In particular, although
recombination affects \emph{which} lineages can coalesce into $j$, it does not
change the \emph{number} of such lineages: in other words, a recombination
event changes $C_j(t)$ but not $I_{j}(t)$.

@@

% [Do we need to get into this above-the-root subtelty here?]
% To ensure that the ancestral material carried by a single lineage can at
% all time be represented by a single half-open interval, we only update the
% endpoints of the involved lineages following common ancestry or recombination events
% such that only the minimal half-open interval whose endpoints have not fully coalesced
% is used to update $L(t)$. Note that empty intervals are not added to the state.
% %By reasoning about the process in this way we can keep representing 
% % all lineages by means of a single half-open interval until $\vert L(t) \vert = 0$
% % and the process terminates.

% %After any coalescence or recombination event,
% %the segment(s) $[x_j,y_j)$ (and $[x_k, y_k]$) carried by the involved lineage(s)
% %are replaced by a single segment $[x, y)$ whose endpoints have not fully
% %coalesced, i.e., $x = \min\{ z \ge \min(x_j, x_k) : N(t,z) > 1\}$,
% %and $y = \sup\{ z \le max(y_j, y_k) : N(t,z) > 1\}$. Note that empty intervals
% %are not added back to $L(t)$.

% Finally, by describing the SMC
% backwards in time, some intervals contained within the half-open
% interval representing a lineage might have fully coalesced before all the genetic
% material carried by that lineage has. As noted by \citet{mcvean_approximating_2005},
% such intervals can be observed as the stalks on the marginal genealogies (e.g.\ marginal trees
% for $[25, 81)$ Fig. \ref{fig:smc-unary}). Recombination events occurring within these smaller
% intervals however do not affect the marginal genealogies.

@@

\begin{figure}
    \centering
    \includegraphics[width=0.65\textwidth]{figures/single_event.png}
    % FIXME commenting this caption out because it mucks up syntax highlighting for me.
    % \caption{
    % Graphical representation of the events associated with the edge $(p,c,x,y)$.
    % The lineage carried by $c$ spans $[x_c,y)$ until 
% a recombination event (blue dotted line) splits it
    %     into two segments. Left and right hand segments coalesce at times $t_p'$ and $t_p$ respectively,
    %     bounding the time to recombination by $(c. \min(t_p, t_{p'})$.
    %     Note that the contribution to the recombination hazard by edge $(p,c,x,y)$ only covers ``half'' of the
    %     green shaded area. The other half is contributed by edge $(p', c, x_c, x)$. 
    %     \label{fig:likelihood}
    % % }
\end{figure}

Now consider the ``hazard'' associated with recombination.
Begin with an edge $e = (p,c,x,y)$.
This implies that there was a sequence of lineages
from the time of the child $c$ (call this time $t_c$)
back to the time of the parent $p$ (again, $t_p$),
that contained the segment $[x,y)$,
and the lineage that directly inherited from $p$
had the segment $[x,y)$ (see Fig.~\ref{fig:likelihood}).
We therefore know there was no recombination between $x$ and $y$
on any of those lineages 
over this time span,
the probability of which is
$\exp(-r \times \mathcal{A})$, where $r$ is the recombination rate
and $\mathcal{A}$ is the total area (eligible links times depth) of the edge.
So, the first contribution to the likelihood from this edge $e$ is
\begin{equation}\label{eq:span}
A_e(r) = \exp(-r (y-x - \mathbbm{1}_{x > x_c})(t_p - t_{c})) .
\end{equation}
Note that if a recombination did occur, then the number of links not
affected by recombination is 1 less (see eq.~\eqref{eq:depth}).

@@

The likelihood of ARG defined by the set of edges $E$ and nodes $N$
given parameters $\theta$ then is
\begin{equation}\label{eq:full-lik}
    \mathcal{L}(\mathcal{G}|r, \lambda)
    =
    (r)^{|E|-|N|+n_r} (\lambda)^{|E|-|N|+n_s} \prod_{(p, c) \in E} A_e(r) B_e(r, \lambda) .
\end{equation}
% One significant advantage of this formulation is that the computation time
% is linear in the number of edges, and an algorithm can be devised to 
% to efficiently keep track of $I_e(t)$ (see Appendix).
% Although the first two factors are easily expressed in terms of numbers of edges and nodes,
% we will see below
% that in practice it is convenient to absorb these terms into per-edge contributions.
% We can also accommodate stepwise changes in the 
% coalescence rate (or recombination rate)
% through time by intersecting the internode intervals with the rate change
% intervals; rate changes along the genome can be incorporated similarly.
Finally, we have only considered the topology and branch length information of any ARG.
Extending this expression by adding the likelihood of observing a set of mutations
given an ARG is given by
\begin{equation}\label{eq:lik-mut}
    \mathcal{L}(\mathcal{D}|\mathcal{G}, \mu)
    =
    \frac{1}{\mathcal{M}!} \prod_{(p, c, x, y) \in E} e^{-(y-x)(t_p-t_c)\mu^{|m_e|}}
\end{equation}
where $\mathcal{M}$ the total number of mutations and $m_e$
the number of mutations along edge $(c,p)$ \citep{mahmoudi_bayesian_2022}.
